Synthesizing Quotient Lenses

### 2017-10-09 papers/QuotientLensSynth/main.tex
Made changes suggested from Friday. Need to add programs and diagrams!

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash} \; (#1, #2, #3)}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}\; (#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}
\newcommand{\canonizer}{\ensuremath{\mathit{Canonizer}}}

@@

\begin{abstract}

\keywords{quotient lenses, quotient regular expressions, synthesis}
\end{abstract}

\section{Introduction}

{\em Lens programming languages} such as
Boomerang~\cite{Boomerang}\bcp{and others!} enable programmers to kill two
birds in one stone by expressing both directions of a bidirectional
transformation with a single program.

\bcp{
  \begin{itemize}
  \item Small example of a boomerang program (maybe in our syntax)
  \item Discussion of lens laws.
  \end{itemize}
}

However, the constraints that are imposed by the lens laws are sometimes too
strict. 
For instance, the programmer may wish to ignore updates to the data that only
change the order of fields, or add whitespace characters that in reality are not
part of the data. Foster et al therefore proposed \textit{quotient lenses} as a
solution to this problem. Quotient lenses enable a programmer to write lenses
which transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

\dpw{The next paragraph gets at the problem, but takes a while to do
so.  Why not make the first sentence go directly to the problem:  ``Unfortunately,
quotient lens programs are unnecessarily verbose and challenging to construct.'' ... or
something like that.}

This paper addresses two main challenges that arise when programming quotient
lenses. The first challenge is tied directly to the programming of quotient
lenses in Boomerang, a string lens programming language where the source and
view types are given by regular expressions. While Boomerang enables the
programmer to express a rich set of quotient lenses, the syntax for writing
these lenses is often too verbose and cumbersome. This makes writing quotient
lenses in Boomerang rather difficult.

For example, suppose that the programmer wishes to convert citation records
stored in \textsc{Bib}\TeX{} format to a representation of the same data in the
EndNote format. The programmer wishes to ``quotient'' away unnecessary
whitespace characters in between the various fields in each record, while also
ignoring the order in which these fields occur in the recore. For instance, the
quotient lens which the programmer wishes to derive should convert this
\textsc{Bib}\TeX{} record

\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
  Publisher = {Printed in GB by William Clowes & Sons Ltd},
  Year = {1971}
}
\end{verbatim}

\noindent to this EndNote record

\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A Conway, J. H.
%D 1971
%I Printed in GB by William Clowes & Sons Ltd
%F conway
\end{verbatim}

\noindent and vice versa.  \bcp{Let's make the example simpler (leave out
  year and publisher?)}

\bcp{Moreover...}
allowing for this equivalent representation of the \textsc{Bib}\TeX{}
entry, which rearranges the fields and uses different whitespace conventions:

\begin{verbatim}
@Book{conway, 
    Publisher = {Printed in GB by 
                 William Clowes & Sons Ltd},

    Title = {Regular Algebra and Finite Machines},

    Year = {1971},

    Author = {Conway, J. H.},
}
\end{verbatim}
\bcp{Can we write (regular and QRE) lenses that handle this version of the
  example?  Write this in Boomerang and see how it looks!!}

While Boomerang enables the programmer to write a lens which exhibits this
behaviour, the syntax tree for the program is large and complicated.

\saz{It might be a good idea to show what this would look like in all its gory
detail, just to make the problem very concrete. Perhaps in a figure?}\bcp{+1}\dpw{+1}

In this paper, we introduce the formalism of Quotient Regular Expressions (or
QREs) to address this issue. \dpw{cut unnecessary phrases that add little information
content.  eg, cut ``the formalism of.''  Doing so makes your text ``punchier.''}  QREs enable the programmer to express
a rich set of equivalence relations on regular expressions at the level of
syntax. QREs therefore enable the programmer to express more equivalence
relations with less effort. Also, since QREs are defined at the level of
syntax on regular expressions, QREs give the programmer more control over where
the quotienting occurs.

\bcp{Give a little example (perhaps for Bibtex and/or EndNote?)}

The second challenge that we address in this paper is the {\em synthesis} of
(bijective) quotient lenses. These are quotient lenses which match the
equivalance classes of the source data to the equivalence classes of the view
data bijectively as in the \textsc{Bib}\TeX{} to EndNote example above. 
Building on our prior work~\cite{popl18} where we showed how to synthesize
(ordinary?) bijective lenses, we show how to synthesize bijective quotient
lenses given a synthesis task described by a pair of QREs and a set of
examples.

Our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a form of quotient lenses
whose types are given by QREs.  Our main technical contribution is a normal
form for QRE lenses and a proof that QRE lenses in normal form are closed
under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

  \subsection{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep R \mapsto s \sep \squash{R}{R'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \;  | \; \normalize{R}{R'}{f}\\
&q' \circ q \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

Each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $EqRel(q)$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $EqRel(q)$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $EqRel(q)$,
  then the equivalence classes of $EqRel(q)$ are the same as the fibres of
  $\canonizer(q)$.

\dpw{I think it would be a little more effective to introduce QREs with a running,
but informal, concrete example. Explain these concepts informally as you go.
Then summarize the syntax at the end.}

  \subsection{Examples}

\bcp{Give some counterexamples too!  And can we take away the enumerate?}

  \begin{enumerate}
    \item 
    Every regular expression can be lifted to a QRE. When a regular expression
    $R$ is used as a QRE, then $\mathcal{L}(R)$ is assumed to be quotiented by
    the equality equivalence relation on $\mathcal{L}(R)$.
    \item
    For the QRE $R \mapsto s$, the data is described by $\mathcal{L}(R)$, but
    all members of $R$ are considered equivalent to each other, with $s \in
    \mathcal{L}(R)$ as the choice of representative. This construct is useful
    when the programmer simply wishes to ignore unnecessary characters such as
    excess whitespace e.g. 
    $$\mathtt{FIRST\_NAME} \cdot (\mathtt{WHITESPACE}^* \mapsto `` \quad ")
    \cdot \mathtt{LAST\_NAME}$$
    \item
    $\squash{R}{R'}{f}$ enables the programmer to use one of two different
    representations of the data if they can provide a way of mapping the first
    representation to the second. For instance, the first representation may be
    $$\mathtt{FIRST\_NAME} \cdot `` \quad " \cdot \mathtt{LAST\_NAME}$$
    while the second representation may be
    $$\mathtt{FIRST\_NAME} \cdot ``," \cdot \mathtt{LAST\_NAME}$$
    The mapping that maps the first representation to the second simply converts
    the space between the first and last name to a comma.
    \item
    The $\perm{q_1, \ldots, q_n}{q}$ construct enables the programmer to
    consider all permutations of some data that are interspersed with a
    separator the same. For instance, in the \textsc{Bib}\TeX{} to EndNote
    example, the programmer may use this construct as follows:
    $$\perm{\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},
    \re{YEAR}}{(\re{WHITESPACE}^* \mapsto \string \n)}$$
 where $\re{REF}, \re{AUTHOR}, \re{TITLE}, \re{PUBLISHER},$ and $\re{YEAR}$ are
regular expressions describing the respective \textsc{Bib}\TeX\; fields of the
same name, and $\re{WHITESPACE}$ is a regular expression describing whitespace
characters.
\item
The constructs $q \cdot q'$, $q \sep q'$ and $q^*$ are similar to the respective
regular expression constructs. However, the regular constructs for QREs are more
restrictive for QREs since the transformations made to the data by
canonizers and lenses involve matching the data against a regular expression
deterministic. Therefore to make these transformations deterministic, the
data should be matched uniquely against the regular expressions, and this
constrains the regular expressions that may be used by QREs. \dpw{This is not an
``example'' -- it is an informal description of how these operators work.}
\item
The $\normalize{R}{R'}{f}$ construct allows the programmer to choose how to
canonize $\mathcal{L}(R)$ into $\mathcal{L}(R')$ be providing the canonizing
function $f$ themselves. This constuct is useful when the desired equivalence
relation cannot easily be expressed by any of the other constructs.
  \end{enumerate}

\saz{Every regular expression $R$ is a QRE.  The main new syntactic
forms are:
$R \mapsto s$, which }

\saz{Would it be simpler to include atoms and $\epsilon$ in $q$ rather than $R$?  That is, $R$ is
  just a subset of $QRE$ built from atoms, $\cdot$ $|$ and ${}^*$.  We could
  reserve $R$ as a metavariable that ranges over $q$'s such that $q$ is a
  regular expression.}

\saz{Need to introduce $\mathcal{L}(R)$ notation.}

\iffalse
Quotient Regular Expressions (or QREs) enable us to give richer specifications
for data formats than ordinary regular expressions. More concretely,
  All data to be transformed must match $W(q)$, the ``whole'' regular
  expression of $q$. However, before transforming the using a lens, the data
  will be quotiented out by the equivalence relation $EqRel(q)$ which has $K(q)$
  as a complete set of representatives. That is, for each string $w \in
  W(q)$, there is a unique $k \in K(q)$ such that $w \; EqRel(q) \; k$). This
  unique representative $k$ is chosen using the ``canonizing'' function
  $\canonizer(q)$.
  \fi

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
As we mentioned earlier, when applying the regular constructs $\cdot
\; | \;  *$ to QREs, we require that the regular languages
described by QREs match the data in only one way so as to make
various transformations made to the data that involve matching the data against
a regular expression deterministic. Consequently, we require that regular
expressions used by QREs be \textit{strongly unambiguous}, a condition which
ensures that if a string $s$ matches a regular expression $R$, then $s$ matches
$R$ uniquely.

@@

The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $EqRel(q)$ and the
canonizing function $\canonizer(q)$ are given in Figures ~\ref{fig:wk},
~\ref{fig:relations} and ~\ref{fig:canonizers}.
\begin{figure}[t]
  \centering
  \[
    \begin{array}{l@{\quad}l@{\quad}l}

@@

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$W(q)^{*!}$}
\AxiomC{$K(q)^{*!}$}
\TrinaryInfC{$q^*$ is well formed}
\end{prooftree}
  \caption{QRE Inference Rules\bcp{Maybe some or all of these can be
      suppressed in the ``short version'' of the paper}\bcp{No, cancel
      that.  These are important.  And we need to comment about the decidability
    issues involving normalize.  (And a note that we don't really use
    normalize, but rather introduce more specialized instances.)}}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended behaviour:
\begin{theorem}
If $q$ is a well formed QRE, then
\begin{enumerate}
  \item $W(q)$ and $K(q)$ are well defined regular expressions,
  \item  $EqRel(q)$ is an equivalence relation on $\mathcal{L}(W(q))$,
  \item  $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $EqRel(q)$,
  \item $\canonizer(q):\mathcal{L}(W(q)) \longrightarrow \mathcal{L}(K(q))$ is a
  well-defined function, and
  \item  given any $w \in \mathcal{L}(W(q))$, $\canonizer(q)(w)$ is the unique
  $k$ in $\mathcal{L}(K(q))$ such that $k$ is equivalent to $w$ modulo
  $EqRel(q)$.
  \end{enumerate}
\end{theorem}

@@

\subsection{Syntax of QRE Lenses}
\subsubsection{Preliminaries}
Given regular expressions $R, S$ and equivalence
relations $\sim_R, \sim_S$ defined on $\mathcal{L}(R)$ and $\mathcal{L}(S)$
respectively, a \textit{bijective quotient lens} $q :
R/{\sim_R}{\Longleftrightarrow} S/{\sim_S}$ from $R$ to $S$ is
a pair of functions $q.get:
\mathcal{L}(R) \longrightarrow \mathcal{L}(S)$ and $q.put : \mathcal{L}(S)
\longrightarrow \mathcal{L}(R)$ such that
\begin{align*}
q.put \; (q.get \; r) &\sim_R r\\
q.get \; (q.put \; s) &\sim_S s
\end{align*}
Additionally the components of $q$ must respect $\sim_R$ and $\sim_S$ i.e.
if $r \sim_R r'$ and $s \sim_S s'$ then $q.get \; r \; \sim_S q.get \; r'$
and $q.put \; s \sim_R q.put \; s' \; s'$.

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\squash}[3]{\ensuremath{\mathit{squash} \; (#1, #2, #3)}}
\newcommand{\perm}[2]{\ensuremath{\mathit{perm}\; (#1)\; \mathit{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\mathit{normalize} \; (#1, #2, #3)}}
\newcommand{\sep}{\ensuremath{\; | \;}}
\newcommand{\canonizer}{\ensuremath{\mathit{Canonizer}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\mathit{get}}}
\newcommand{\lput}{\ensuremath{\mathit{put}}}
\newcommand{\create}{\ensuremath{\mathit{create}}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}

@@

\begin{abstract}
{\em A Quotient lens} is an expression which encapsulates a bidirectional
transformation whereby each of the two transformations that are part of the lens
have the same behaviour on data that are equivalent to each other modulo some
equivalence relation. Quotient lenses save the programmer the effort of writing
two expressions instead of one, give guarantess on how the forward and backward
transformations interact with each other, and also allow the programmer to
ignore differences in the data which are not intended to affect the
behaviour of the transformations.

However, quotient lenses are sometimes difficult to program because the
equivalence relations defined on the data, as well as the underlying lens which
operates on the equivalence classes are complicated. These two challenges make
quotient languages impractical to use in many cases spite of their advantages.

In this paper, we address both of these challenges. First, we propose the
language of {\em Quotient Regular Expressions} or QREs, which enables the
programmer to explicitly define a broad class of equivalence relations on
regular languages using a compact and convenient notation. We then define a
class of {\em QRE lenses}, a class of quotient lenses which are typed using
QREs, and show how to {\em synthesize} QRE lenses from their QRE types and a
set of input-output examples. Finally, we demonstrate the applicability of this
approach by synthesizing QRE lenses between several real-world data formats
from the {\tt data.gov} database.
\end{abstract}

\section{Introduction}
{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression. An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

\noindent stored in the \bibtex format to their equivalent representation 
\begin{verbatim}
%0 Book
%T Regular Algebra and Finite Machines
%A Conway, J. H.
%F conway
\end{verbatim}

in the EndNote format, and vice versa.

Bidirectional programs have also been called {\em lenses}. Formally, Foster et
al~\cite{foster} defined a lens $\ell$ that maps between an ``abstract" {\em
source} $S$ and a ``concrete" {\em view} $V$ to be a triple of functions
\begin{align*}
\ell.get &: S \longrightarrow V\\
\ell.put &: V \longrightarrow S \longrightarrow S\\
\ell.create &: S \longrightarrow V
\end{align*}

\noindent satisfying the equations
\begin{align*}
\ell.\lput \; (\ell.\get \; s) \; s &= s\\
\ell.\get \; (\ell.\lput \; v \; s) &= v\\
\ell.\lput \; (\ell.\create \; v) &= v\\
\end{align*}

The $\get$ function presents a view of the data from the source, the
$\lput$ function enables an updated view to be folded back into the source, and
the $\create$ component produces a copy of the source from a view, using
defaults to fill in information that may have been discarded by the $\get$
function in producing the view. The first law states that the $\lput$ function
must restore all the information discarded by $\get$ when its arguments are an
abstract structure and a concrete structure that generates the very same
abstract structure. The second and third laws state that $\lput$ and $\create$
must propagate all of the information contained in their abstract arguments to
the concrete structure they produce. These laws express fundamental expectations
about how the components of a lens should work together.

However, the constraints that are imposed by the lens laws are sometimes too
strict. For instance, the programmer may wish to ignore updates to the data
that only change the order of fields, or add whitespace characters that in reality are not
part of the data. Foster et al therefore proposed \textit{quotient lenses}
~\cite{quotient} as a solution to this problem. Quotient lenses enable a
programmer to write lenses which transform data that is quotiented out by an
equivalence relation, so that records which are equivalent to each other modulo
this relation are considered the same by the underlying lens.

Unfortunately, quotient lenses are sometimes difficult to program because the
equivalence relations defined on the data, as well as the underlying lens which
operates on the equivalence classes are complicated. For example, suppose that
the programmer wants to program the \bibtex to EndNote transformation in such a
way that the program allows for any of two representations of the authors
name (e.g. ``Conway, J. H." and ``Conway J. H"), ignores the order in which the
author and title fields occur in each record, and allows for arbitrary amounts
of whitespace in between the author and title fields. These rules allow for
these two \bibtex citation records to be considered the same:

\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}

@Book{conway,
    Title = {Regular Algebra and Finite Machines},

    Author = {Conway J. H.},
}
\end{verbatim}

The Boomerang quotient lens which does the \bibtex to EndNote conversion is

\sam{TODO: PUT BOOMERANG EXPRESSION HERE!},

an expression which is difficult to write and understand.

We address the challenge of programming quotient lenses in two ways. First, we
propose the language of {\em Quotient Regular Expressions} or QREs, which
enables the programmer to explicitly define a broad class of equivalence
relations on regular languages using a compact and convenient notation. For
example, the QRE which describes the equivalence relation on the \bibtex
citation records is as follows:

\sam{TODO: PUT QRE FOR BIBTEX HERE!}

This expression better describes the equivalence relation since it is explicit
and more precise.

However, even though QREs enable the programmer to describe complicated
equivalence relations on data more simply, the underlying lens which maps the
equivalence classes of the \bibtex records to the equivalence classes of the
EndNote records is also challenging to program, mostly because of the verbosity
of the expression:

\sam{TODO: PUT QRE LENS FOR BIBTEX HERE!}

Therefore, to address this challenge we demonstrate how to {\em synthesize}
quotient lenses from a pair of QREs and a set of input-output example pairs. Our
key insight is to reduce the problem of synthesizing QRE lenses to the problem
of synthesizing {\em bijective lenses}, a problem which we solved in out prior
work ~\cite{popl18}.

Our approach is to define a class of {\em QRE lenses} which are the same as
bijective lenses with canonizers at the ends. That is, a QRE lens uses the
source QRE to compute a unique representative for a datum modulo the
equivalence relation defined by the QRE before applying a bijective lens to
this representative in the forward direction, and operates similarly same in the
backward direction but using the view QRE.

\sam{TODO: PUT LENS WITH CANONIZERS ON THE EDGE DIAGRAM HERE!}

One potential pitfall with QRE lenses is that the composition of QRE lenses may
not have the normal form of a bijective lens with canonizers on the ends:

\sam{TODO: PUT COMPOSITION DIAGRAM HERE!}

Our main technical contribution in this paper is a proof that QRE lenses are
closed under left quotienting, right quotienting, composition, and regular
combinators. This proof will allow justify our approach for synthesizing
quotient lenses so that the programmer can simply type

$$\mathit{synth} \; \mathtt{BIBTEX} \; \mathtt{ENDNOTE}$$

to derive a quotient lens that does the \bibtex to EndNote.

In summary, our main contributions are:
\begin{enumerate}
\item We introduce a novel language of {\em Quotient Regular Expressions}
(QREs) that provide compact, convenient notation for a natural and useful
class of equivalence relations on regular languages (Section~\ref{QRE}).
\item We define a language of {\em QRE lenses}, a form of quotient lenses
whose types are given by QREs.  Our main technical contribution is a normal
form for QRE lenses and a proof that QRE lenses in normal form are closed
under left quotienting, right quotienting, composition, and regular
operators (Section~\ref{QRE-lenses}).
\item Using this normal form, we reduce the problem of {\em synthesizing}
QRE lenses from their QRE types and a set of examples to the previously studied
problem of synthesizing bijective lenses between regular languages 
(Section~\ref{synth}).
\item We extend the Boomerang {\em implementation} with QREs, QRE lens operators,
and QRE lens synthesis and demonstrate its applicability by using it to
synthesize QRE lenses between several real-world data formats from the
{\tt data.gov} database (Section~\ref{impl}).
\end{enumerate}
Sections~\ref{relwork} and~\ref{concl} discuss related and future work.

@@

For example, suppose that the programmer wishes to write a regular expression
which describes \bibtex citation records such as:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

Each record may be represented in many different ways. Firstly, the author's
name can have the last name followed by a comma, and then the
first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. For example, each of these
\bibtex records is equivalent to the other:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Author = {Conway J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}

Secondly, the author and title fields may occur in either order:
\begin{verbatim}
@Book {conway,
  Author = {Conway, J. H.},
  Title = {Regular Algebra and Finite Machines},
}
\end{verbatim}
\begin{verbatim}
@Book {conway,
  Title = {Regular Algebra and Finite Machines},
  Author = {Conway, J. H.},
}
\end{verbatim}
Thirdly, the title and author records can be separated by arbitrary amounts of
whitespace characters. For example, if the title field occurs first, then the
author field may occur not on the line immediately below the title field, but
instead on the second line below the title field:
\begin{verbatim}
@Book {conway,
  Title = {Regular Algebra and Finite Machines},

  Author = {Conway, J. H.},
}
\end{verbatim}
In the example above, a \bibtex record may have different
representations for three different reasons. The first reason
is that the author's name can have the last name followed by a comma and then
the first and second names separated by a space, or the last name, first name and
second name separated by spaces in between them. The $\squash{R}{R'}{f}$ QRE
enables the programmer to choose between the representations described by $R$
and $R$', provided that they can provide a function $f$ that maps data stored in
the first format to data stored in the second format. In this example we have
the two regular expressions

\begin{align*}
\mathtt{COMMA\_NAME} &= \mathtt{NAME} \cdot ``," \cdot \mathtt{NAME} \cdot ``
\quad " \cdot \mathtt{NAME}\\
\mathtt{SPACE\_NAME} &= \mathtt{NAME} \cdot ``," \cdot \mathtt{NAME} \cdot ``
\quad " \cdot \mathtt{NAME}\\
\end{align*}

and the bijective lens 
$$ \ell = (\mathit{copy} \; \mathtt{COMMA\_NAME}) \cdot (\mathit{ins} \; ``,")
\cdot (\mathit{del} \; `` \quad") \cdot (\mathit{copy} \; \mathtt{COMMA\_NAME})
\cdot (\mathit{copy} \; \mathtt{COMMA\_NAME})$$
therefore the QRE
$$\mathtt{AUTHOR}
=
``\text{Author
= \{}" \cdot \squash{\mathtt{COMMA\_NAME}}{\mathtt{SPACE\_NAME}}{\ell.get}
\cdot ``\}"$$ enables the programmer to express the fact that the two different
representations of the author's name are equivalent, with the
$\mathtt{COMMA\_NAME}$ representation chosen as the canonical representation,
and the function $\ell.get$ giving the ``canonizing'' function which converts
data in the $\mathtt{SPACE\_NAME}$ format to its $\mathtt{COMMA\_NAME}$
representation. The $\squash{R}{R'}{f}$ has the added advantage that if the
transformation from $R$ to $R'$ can be expressed by a bijective lens,
then the programmer may attempt to \textit{synthesize} the lens from $R$ to $R'$
because of our previous work where we demonstrated how to synthesize bijective
lenses from a pair of regular expressions and a set of input-output example
pairs:
$$\mathtt{AUTHOR}
=
``\text{Author
= \{}" \cdot \squash{\mathtt{COMMA\_NAME}}{\mathtt{SPACE\_NAME}}{\mathit{synth}
\; \mathtt{COMMA\_NAME} \; \mathtt{SPACE\_NAME}} \cdot ``\}"$$

\sam{Should we have a special construct like squash-synth(R, R') in addition to
squash(R, R', l) or maybe even just have only squash-synth(R, R')? Perhaps
having to explicitly write the lens weakens our case that QREs are compact and
convenient}

The second reason that a record may have two different representations is that
the author and title fields may occur in a different order, and the third
reason is that the author and title fields may have arbitrary occurences of
whitespace characters in between them. The $R \mapsto s$ and $\perm{q_1,
\ldots, q_n}{q}$ QREs enable the programmer to handle each of these cases:
$$\perm{\re{AUTHOR}, \re{TITLE}}{((``," \cdot \re{WHITESPACE}^*) \mapsto \string
``, \backslash n")}$$
This QRE enables the programmer to consider any permutation of the author and
title fields interspersed with a comma and arbitrary whitespace as equivalent.
The $((``," \cdot \re{WHITESPACE}^*) \mapsto \string ``, \backslash n")$
subexpression is itself a QRE that denotes that all string $s$ which consists
of a comma $``,"$ followed by any number of whitespace should be considered
equivalent, with the canonical representative being the string $\string ``,
\backslash n"$. The canonical representation of \bibtex records is that in
which the author field occurs the title field with a comma then a newline
character separating the two.

The regular combinators for QREs i.e. $\cdot$ (concatenation), $|$ (union) and
$*$ (iteration) behave like the regular combinators for regular languages. For
instance, if the QRE for matching a \bibtex \; record is named
$\mathtt{RECORD}$, then the QRE for matching two records is $\mathtt{RECORD}
\cdot \mathtt{RECORD}$, the QRE for matching one or two records is $\mathtt{RECORD}
\sep (\mathtt{RECORD} \cdot \mathtt{RECORD})$ and the QRE for matching any
number of QRE records in $\mathtt{RECORD}^*$.

The $\normalize{R}{R'}{f}$ QRE is the most general QRE in that for each of the
the other QREs $q$, there exist regular expressions $R, R'$ and a surjective,
idempotent function $f:\mathcal{L}(R) \longrightarrow \mathcal{L}(R')$ such that
$q$ $\normalize{R}{R'}{f}$ each denote the same equivalence relation on the
same regular language. $\normalize{R}{R'}{f}$ enables the programmer to
provide a canonizing function from $\mathcal{L}(R)$ to $\mathcal{L}(R')$,
with the equivalence class on $\mathcal{L}(R)$ being the equivalence relation
defined by the {\em fibres} of $f$ (i.e. $r \sim r'$ if and only if $f(r) =
f(r'))$. The canonizing function is required to be surjective and idempotent.
The $\normalize{R}{R'}{f}$ combinator therefore enables the programmer to express
equivalence relations which are difficult to express, or which cannot be
expressed using the other combinators.

In summary each QRE $q$ enables us to express
\begin{enumerate}
  \item a regular expression $W(q)$ (the ``whole'' of $q$),
  \item an equivalence relation $\eqrel{q}$ on $\mathcal{L}(W(q))$,
  \item a regular expression $K(q)$ (the ``kernel'' of $q$)
  such that $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$, and
  \item a ``canonizing'' function $\canonizer(q):\mathcal{L}(W(q))
  \longrightarrow \mathcal{L}(K(q))$ which given any $w \in \mathcal{L}(W(q))$,
  computes $\canonizer(q)(w)$ as the unique $k$ in $\mathcal{L}(K(q))$ such that
  $k$ is equivalent to $w$ mod $\eqrel{q}$
  \end{enumerate}
  Observe that since $\canonizer(q)$ computes $\canonizer(q)(w)$ as the unique
  $k$ in $\mathcal{L}(K(q))$ such that $w$ is equivalent to $k$ mod $\eqrel{q}$,
  then the equivalence classes of $\eqrel{q}$ are the same as the fibres of
  $\canonizer(q)$.
\subsection{Syntax of QREs}
The language of Quotient Regular Expressions (QREs) is given by the following
grammar:
\begin{align*}
q := \; &R \sep R \mapsto s \sep \squash{R}{R'}{f} \sep
\perm{q_1, \ldots, q_n}{q} \;  | \; \normalize{R}{R'}{f}\\
&q' \circ q \sep q \cdot q' \sep (q \sep q') \sep q^*,
\end{align*}
where $R$ ranges over regular expressions, $f$ ranges over functions between
regular languages, and $s$ ranges over character strings.

\subsection{Semantics of QREs}
\subsubsection{Preliminaries}
Let $q, q'$ be QREs. When applying the regular combinators to QREs, we require
that if a string $s$ matches any of the regular expressions $W(q) \cdot W(q')$,
$W(q) \sep W(q')$, $W(q)^*$, $K(q) \cdot K(q')$,
$K(q) \sep K(q')$, $K(q)^*$, then $s$ matches that regular expression in
only one way. This unambiguity condition is called {\em strong unambiguity},
and is necessary, firstly because it ensures that the canonizing function of a
QRE is well-defined. For example if $\mathtt{RECORD}$ is the QRE for a \bibtex
record in our \bibtex to and $W(\mathtt{ENDNOTE} \cdot \mathtt{ENDNOTE})$ is not
unambiguous, that is there is a string $s$ such that $s = s_1 \cdot s_2 =
{s_1}' \cdot {s_2}'$ with $s_1, s_2, {s_1}', {s_2}' \in W(\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE})$, then the canonizer $f$ for $\mathtt{ENDNOTE} \cdot
\mathtt{ENDNOTE}$ may not be well defined since we are not guaranteed that
$f(s_1) \cdot f(s_2) = f({s_1}') \cdot f({s_2}')$. We also require that the
regular combinators applied to the kernels are unambiguous since the underlying
lenses will end up operating on the kernels, and these lenses impose the same
restrictions for similar reasons.

@@

The semantics for QREs that formally define the whole language $W(q)$, the
kernel language $K(q)$, the equivalence relation $\eqrel{q}$ and the
canonizing function $\canonizer(q)$ are given in
Figures~\ref{fig:wk},~\ref{fig:relations} and~\ref{fig:canonizers}.
\begin{figure}[t]
  \centering
  \[
    \begin{array}{l@{\quad}l@{\quad}l}

@@

\begin{prooftree}
\AxiomC{$q$ is well formed}
\AxiomC{$W(q)^{*!}$}
\AxiomC{$K(q)^{*!}$}
\TrinaryInfC{$q^*$ is well formed}
\end{prooftree}
  \caption{QRE Inference Rules\bcp{Maybe some or all of these can be
      suppressed in the ``short version'' of the paper}\bcp{No, cancel
      that.  These are important.  And we need to comment about the decidability
    issues involving normalize.  (And a note that we don't really use
    normalize, but rather introduce more specialized instances.)}}
  \label{fig:qrerules}
\end{figure}
The theorem below confirms that the semantics of QREs are consistent with their
intended behaviour:
\begin{theorem}
If $q$ is a well formed QRE, then
\begin{enumerate}
  \item $W(q)$ and $K(q)$ are well defined regular expressions,
  \item  $\eqrel{q}$ is an equivalence relation on $\mathcal{L}(W(q))$,
  \item  $\mathcal{L}(K(q))$ forms a complete set of representatives for
  $\eqrel{q}$,
  \item $\canonizer(q):\mathcal{L}(W(q)) \longrightarrow \mathcal{L}(K(q))$ is a
  well-defined function, and
  \item  given any $w \in \mathcal{L}(W(q))$, $\canonizer(q)(w)$ is the unique
  $k$ in $\mathcal{L}(K(q))$ such that $k$ is equivalent to $w$ modulo
  $\eqrel{q}$.
  \end{enumerate}
\end{theorem}

@@

Our approach in defining QRE lenses is as follows. Let $c, c'$ be QREs. Given a
bijection $\ell : K(c) \Leftrightarrow K(c')$, we may define a quotient lens $q
: W(c)/\eqrel{c} \Leftrightarrow W(c')/\eqrel{c'}$ by
\begin{equation}\label{normalform}
q.get = \ell \circ \canonizer(c) \text{ and } q.put = \ell^{-1} \circ
\canonizer(c'),
\end{equation}

A potential problem with this approach is that quotient lenses of this form may
not necessarily be closed under the action of the lens combinators. For
example, given two lenses $q$ and $q'$, the composition of $q' \circ q$ may not
be of this form.
